\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-19 coqimpl.tex
Adding Language Design Section.

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. To achieve incremental type checking, we notice the nature of late binding: for each defining field $ .. \cL\sigma \vdash t : T $, we can compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we use module and module parameters to achieve this wrapping---each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each field of the family will compile into a parametrized module (functor). Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} Then, these modules and module types will be generated and type-checked by Coq immediately. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding---for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list.

@@

An inheritance data will be annotated with the type for the input family and output family, and generally there are three kinds of inheritance---extend, override and inherit. Extending corresponds to adding new fields, overriding corresponds to modify existent fields and inheriting corresponds to directly taking fields from the parent. Thus it is possible to consider inheritance as solely a data rather than a functor, and this actually provides us the ability to \text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.}
Unlike general-purposed programming, Coq should be able to  \textit{reason} about each program. However, sometimes reasoning requires a specific program and overriding on them is unacceptable. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in \cref{fig:plugin-example1}, consider we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more.  To fix that, we need more information on "a" than its type---we need to expose its concrete definition "a = 1" to appear in the context of "b".

To achieve such exposure of definition, in meta-theory we only need to expose the information as propositional equality as part of the signature, and in our Coq plugin, we can achieve that even easier by exposing the whole definition directly into module type. Doing so we can have a better judgemental equality and we can avoid overriding because subtraction doesn't have the same definition as $add$. We require the users to decorate field as `Final' so that the plugin will proceed with this special treatment.

\textbf{Special Fields: Overridable and Sealed Family as Interface.}
But there are cases reasoning and overriding are both required, especially when we need to use Coq to do computation. For example, we may want a field to be an arbitrary monotonic function. Again, we cannot split into two fields \mintinline{Coq}{(f : nat → nat, p : monotonic f)} because to provide "p" we need to have concrete definition of "f" in the family type, which makes "f" non-overridable at all. One option is to use sigma type \mintinline{Coq}{{f | monotonic f}} but that is cumbersome when we have a bundle of stuff with multiple refinements.

To resolve this, we provide family sealing mechanism---basically we allow the users to specify the family/module type, and thus we can have a ``local'' transparent, non-overridable field located in a overridable family. Using the above example, we simply make \mintinline{Coq}{(f : nat → nat, p : monotonic f)} together inside an overridable family, but "f" to be non-overridable and transparent on its definition, and thus we can prove "p". Once we want to swap the implementation, we just override the whole ``sealed'' family.

This ``sealed'' family actually simulate interface to a certain level---we can assign each field of the sealed family with \mintinline{Coq}{Axiom non_implement : forall {T : Type}, T.}  , and later instantiate them with concrete implementation by overriding this family.
% Basically, {a : int -> int; property : a is monotonic}, in this case
%   we may want to override "a" using different computation, but we still wnat
%   this refinement/property

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive type in Coq (because that require non-trivial research effort on theory side---i.e. {What is the semantic of extensible inductive type?}---and engineering), but we simulate extensible inductive type using overriding---when extending an inductive type $A$ with new constructors $c$, our plugin will generate a new inductive definition $A'$ with this new $c$ on the fly and feed to Coq.  But still, we must be careful on the exposing data in the family type (module type) for the following field.

@@

But we do have to provide a way to construct a recursor, and execute type checking (or exhaustiveness checking). Thus we need to extract a elimination principle for future usage. Here we show one exemplar extraction of the recursor in \mintinline{Coq}{Set} as module "b_rec_12".

To construct a recursor, we make for each constructor a handler and aggregate them inside a family. The whole family "neg_handler" can be considered as a large matching case. Doing so, family inheritance can function as matching-case reuse. Then we use "FRecursor" to specify the motive and inductive type. Here "FRecursor" will feed Coq
a module "v_33_34" for doing type-checking---what it does is basically assert the existence of the recursor and apply to see if handlers can pass the type checking by Coq.

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. We may have to alter the formulation of the recursor.

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
However, to verify and prove the recursor and partial recursor (and their computational rules), our plugin has to compile the extensible family into a non-extensible Coq Module and prove the recursors by instantiating the content of Coq module with concrete inductive type.

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'', including the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning---we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus not easy to handle.

This "Closing Fact" can ease us from the necessity to let plugin generate proof for recursors and partial recursors, and thus making plugin development easier. For example, the computation rule for a concrete recursor like "B.neg" and partial recursors in \cref{fig:plugin-example2} can be directly asserted by using "Closing Fact" and verified later during compilation. This ``Global Reasoning'' will be again useful in the later examples.

>>>>

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. Unlike module, we need to faithfully reflect the \textit{late binding} nature of family : for each defining field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we choose module and module parameters to achieve this wrapping---each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each field of the family will compile into a functor parametrized by its context. Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} Then, these modules and module types will be generated and type-checked by Coq immediately. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding---for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list. This compiled module type serve as an abstraction to let overriding possible.

@@

An inheritance data will be annotated with the type for the input family and output family, and generally there are three kinds of inheritance---extend, override and inherit. Extending corresponds to adding new fields, overriding corresponds to modify existent fields and inheriting corresponds to directly taking fields from the parent. Thus it is possible to consider inheritance as solely a data rather than a functor, and this actually provides us the possibility to \text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy can allow fields to have type dependent on former fields thus fulfilling \ref{langdesign-req1}. However, there are some caveats, especially when the form fields are actually ``proof-relevant''. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in \cref{fig:plugin-example1}, consider we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more (and this is necessary so that "a" can be overridden in the future). To fix that, we need more information on "a" than its type---we need to expose its concrete definition "a = 1" to appear in the context of "b".

To achieve such exposure of definition, in meta-theory we only need to expose the information as propositional equality as part of the signature, and in our Coq plugin, we can achieve that even easier by exposing the whole definition directly into module type. Doing so we can have a better judgemental equality and we prohibit overriding because subtraction doesn't have the same definition as $add$. We require the users to decorate field as `Final' so that the plugin will proceed with this special treatment.

\textbf{Special Fields: Overridable and Sealed Family as Interface.} 
Non-overridable field is still not expressive enough to achieve the full extent of the dependent type programming. There are cases reasoning and overriding are both required, especially when we need to use Coq to do computation. For example, we may want a field to be an arbitrary monotonic function. Again, we cannot split into two fields \mintinline{Coq}{(f : nat → nat; p : monotonic f)} because to provide "p" we need to have concrete definition of "f" in the family type, which makes "f" non-overridable at all. One option is to use sigma type \mintinline{Coq}{{f | monotonic f}} but that is cumbersome when we have a bundle of stuff with multiple refinements.

To resolve this, we provide family sealing mechanism---basically we allow the users to specify the family/module type, and thus we can have a \textit{locally transparent}, \textit{non-overridable} field located in a \textit{overridable} family. Using the above example, we simply make \mintinline{Coq}{(f : nat → nat, p : monotonic f)} together inside an \textit{overridable} family, but "f" to be non-overridable and transparent on its definition, and thus we can prove "p". Once we want to swap the implementation, we just override the whole ``sealed'' family.

This ``sealed'' family actually simulate interface to a certain level---we can assign each field of the sealed family with \mintinline{Coq}{Axiom non_implement : forall {T : Type}, T.}, and later instantiate them with concrete implementation by overriding this family.
% Basically, {a : int -> int; property : a is monotonic}, in this case
%   we may want to override "a" using different computation, but we still wnat
%   this refinement/property

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive type from \ref{langdesign-req2} in Coq (because that require non-trivial research effort on theory side---i.e. {What is the semantic of extensible inductive type?}---and engineering), but we simulate extensible inductive type using overriding---when extending an inductive type $A$ with new constructors $c$, our plugin will generate a new inductive definition $A'$ with this new $c$ on the fly and feed to Coq.  But still, we must be careful on the exposing data in the family type (module type) for the following field.

@@

But we do have to provide a way to construct an (extensible) recursor (as in \ref{langdesign-req3}), and execute type checking (and exhaustiveness checking). Thus we need to extract a elimination principle for future usage. Here we show one exemplar extraction of the recursor in \mintinline{Coq}{Set} as module "b_rec_12".

To construct a recursor as in \ref{langdesign-req3}, we make for each constructor a handler and aggregate them inside a family. The whole family "neg_handler" can be considered as a large matching case. Doing so, family inheritance can function as matching-case reuse. Then we use "FRecursor" to specify the motive and inductive type. Here "FRecursor" will feed Coq
a module "v_33_34" for doing type-checking---what it does is basically assert the existence of the recursor and apply to see if handlers can pass the type checking by Coq.

@@

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref{fig:plugin-example2} is not expressive enough to really fulfill \ref{langdesign-req2}. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. Thus, apart from the (total) recursor, we need to have another formulation of the recursor i.e., we need to have a better characterization of ``extensible inductive type'' exported in the module type as interface.

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
However, to verify and prove the recursor and partial recursor and their computational rules (as in \ref{langdesign-req4}), our plugin has to compile the extensible family into a non-extensible Coq Module and prove the recursors by instantiating the content of Coq module with concrete inductive type.

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'', especially the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning---we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus a bit harder to handle.

This "Closing Fact" can ease us from the necessity to let plugin generate proof for recursors and partial recursors and their computational rules (as in \ref{langdesign-req4}), and thus making plugin development easier. For example, the computation rule for a concrete recursor like "B.neg" and partial recursors in \cref{fig:plugin-example2} can be directly asserted by using "Closing Fact" and verified later during compilation. This ``Global Reasoning'' will be again useful in the later examples.

\textbf{Hack: ``Meta-data''.}

### 2022-06-19 lang-design.tex
Adding Language Design Section.

When adapting Family Polymorphism into dependent type, we choose to shrink and focus on only the essence of family (and inheritance) structure in \citet{zm2017}, and thus a lot of unrelated feature, like Interface, will be tailored. In this case, it will look like module with late-binding and inheritance as code reuse for module.

We also need it to have a good compatibility with Inductive Type, because we don't want to lose too much reasoning power of Coq.

>>>>

When adapting Family Polymorphism into dependent type, we choose to tailor and focus on only the essence of family (and inheritance) structure in \citet{zm2017}, and thus a lot of unrelated feature, like Interface, will be removed. In this case, it will look like module with late-binding and inheritance as code and proof reuse for module. We also need it to have a good compatibility with Inductive Type, because we don't want to lose too much reasoning power of Coq (and its tactic programming).

Let's start with the basic example---STLC and its extension---to consider what kind of features are required and how much of them can be supported by family polymorphism.
% add the example code here

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to reason about fields in a family}, just like in a module, where a field can reason about its former fields. More generally, we want it possible for a later field to have type dependent on the former fields.

% we want extensible inductive type
\textlabel{Req (2)}{langdesign-req2}:\textbf{ we want extensible inductive type}. The extension of inductive type is not supported by \citet{zm2017} thus we need further consideration about it on both implementation and metatheory. Additionally, \textlabel{Req (3)}{langdesign-req3}:\textbf{we also want extensible pattern matching and induction reasoning}. There are actually two kinds of ``pattern matching'', one for data and the other for induction reasoning, i.e. one use the eliminator to \mintinline{Coq}{Set} or \mintinline{Coq}{Type}, and the other uses the eliminator to \mintinline{Coq}{Prop}. Luckily, in this setting, extensible pattern matching can be easily expressed with family extension -- we just aggregate all case handlers of pattern matching into a family (as a bunch of functions), and then one family can encode one pattern matching, and family inheritance can express adding case handlers. Then we just need to introduce a primitive that will ``wrap'' that family into a recursive function. Induction reasoning can be handled in an almost identical way.

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction reasoning, because the later one is \textit{proof-irrelevant}. This difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:\textbf{we want to reason aboue extensible pattern matching}. Our substitution lemma is one example,

% we want tactic programming and certain level of automation
And finally, \textlabel{Req (5)}{langdesign-req5}:\textbf{we want tactic programming}.
